Discussion Draft

### 2020-03-31 doc/Discussion.rmd
incorporating thoughts from new discussion outline

---
title: "Discussion Draft"
date: ""
output:
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document: default
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
bibliography: refs_Main_Report.bib
---

@@

@@

The 'typical landscapes' refers to the sex differences in the board scale pattern or general placement of crossovers along the length of chromosomes. The primary pattern which has been observed across mny species is that crossovers are shifted to telomereic positions in males while a more unifrom landscape in observed in females [@sardell_sex_2020].

Our results show that the typical landscape is consered across all strains of house mice in this study (for all single chromosome daata and when controlled for crossover number), the broad scale position of crossovers was shifted closer to the telomere end (of the long arm).

<Detangling centromere suppression from telomere bias,  The Sardel guppies paper was able to deduce (dientangle) that this typical landscape pattern was due to // disentangle the specific components of centromere supression and telomere bias between the male and feamel landscapes. we are not able to do that in these cases (because of inability to estimate the chromosome identity across our sample of single bivalent
s. )>

The conclusions from the models -- basically all of the models could account for differences in the general recombination landscape patterns.

@@

**the within animal variation is equvillent to between cell variation**

Our results confirm previous findings from mice [@lynn2002] and humans of females having greater between cell variation compared to males.

Transcriptional noise and complex traits both show the same repsonses to different modes of selection, lower variation under stabilizing selection and greater variaion with relaxed selection. The difference in between

so all of the models could reasonably predict the above results ... examining some of the traits related to the faster evolution in the male specific data indecate different support across the models.

None of the models make predictions for between cell variation -- so we draw conclusions from their preciations based on the strength of interference.

# Chromatin Structure and interference strength

@@

@@

Another thing to point out / regarding this model -- is the connection to sex difference in interference strength. Our results of conserved within strain raw IFD lengths - yet when the IFDs are controlled for total SC length, we observed sexual dimorphism with males having longer IFD^normalized^ (stronger interference), mirror those observed in [@petkov2007].
The model / explanation proposed in that paper is that The chromatin structure which has longer loops (and shorter axis) will have more DNA bps folded / sequestered into the DNA loops. Thus the strength of interference will be stronger because with each step (count) along the 2D axis, more bases pairs are being skipped above the axis. These findings point out two things, This part of heterochiasmy generates testable prediction: First, the interference strength measured in chromosome axis units (or SC) is generally conserved between sexes. Second the sex with longer axis and shorter loops will have weaker interference compared to the other sex. Or conversely, the sex with stronger linkage map based interference would have chromatin organization with shorter axis compared to the other sex. There is a plethora of data of sexual dimorphism in linkage map based interference strength indicating sexual dimorphism, but comparative cytological data is lacking (difference in perparness of some sexes and complications due to identification of specific chromosomes between cells.)

@@

# Revered heterochiasmy

```{r reverse.HetC, echo=FALSE}
# transition connection to the hetC models

#our results - total gwRR isn't the most conserved, (we observed 3 other traits more)

#what does a reversal of HetC imply?  gwRR
#driven by faster evolution in males

#our model had faster evolution in males resulting in a reversal of the canonical pattern of heterichiasmy ( F gwRR > M gwRR)
#other examples, cattle, marsupials

#what do models say about faster evolution in one sex? (yes -- but almost always for canonical pattern of heterochiasmy)

Our results demonstrate that gwRR is decoupled from other sexually dimorphic meotic traits, (we observed 3 other traits more conserved sexual dimorphism examined above). We the general pattern 2 strains attribute this pattern of reversed heterochiasmy to faster evolution of gwRR in two strains males.

Which models predict different rates of evolution between the sexes. Some of the models could explain a reversal of the direction of sexual dimrohpsim, however most all of the predictions for faster evolution result in males having reduced recombination rates. Under the haploid selection model a  **reversal of the direction of heterochiasmy** would be accompanied by vast changes in haploid stages or degree of imprinting. Following the meiotic drive based model, females should be predicted to have larger linkage maps/gwRR since they are more exposed / weaker to true meiotic drivers. A reversal in heteroschiasmy is not clearly predicted by the SACE model (unless with the -- except under the conditions described below). Under the spindle selection model, the male (centrosome/effect SAC) metaphase could drive directional selection (and faster evolution) in males compared to female. However the mechanism/aspect of different fitness is not clear. Under the COM model, evolutionary predictions are not clear (explicit), but following the logic would come from changes in axis length (to generate another wave point) -- which might be due to meiocyte volume.

@@

@@

Our results indicated that interference strength has evolved with the genome wide recombination rates in the three strains with faster male evolution.

This is unexpected because most literature and logical models are built / assume that weaker dense spacing of crossovers along chromosomes (ie. weaker interference) is how. The history, models and empirical examples are extensively reviewed in [@ottoPaysuer2019]. Broad scale patterns for the correlation of gwRR and interference strength (gamma parameter) do fit the expected negative pattern [@ottoPaysuer2019].

There's no model that fits evolution of heterochiasmy and interference together. Under our undertsanding of the models, the relatioship of genome wide recombiantion rate and interference strength is assumed to be negative. We now review how some of the assumptions of typical modifer models might not fit realities of the meiotic pathway.

First, the **typical modifer models do not model single meiotic events**. We conclude that under the haploid selection and meiotic driver model any prediction for evolution of interference strength follows from the 'logical' expectation of a negative correlation.

This model predicts the evolution of heterchiasmy -through models (that build on selection to reduce recombiantion for the sex with geater varatince in fitness effect)

**SACE**
While the SACE model doesn't explecitly make predictions aof the out come for interference strength (or averge spacing of crossovers)
We note that the main predicted pattern under this model of maintaining larger blocks of chromosomes in linkage can asrise from stronger interference / greater spacing of the distance between two foci (also maximize rbar).

Under the **COM** model -- the placement of crossovers (the recombination landscape) is generated by 'waves' / movement from the bouquet stage.  It is conceivable that the pattern of stronger interference could be generated from the waves/movements of chrmosomes in the early prophase stage this would come from a change in the movement or extension of the axis length.

The results from golstein reviewed in [@ottoPaysuer2019], predict a modeifier model (which placed more weight on crossover assurance (holds the number of crossovers constant) and simulates a scnario where a modifer acts on modulating the probabilty of flanking regions having 0 or 2 crossovers. Which has the result of increasing the distance between two foci.

Additionally there are a handful of empircal results which also show the unexpected possitive correlation between gwRR and interference strength. When comparisons of genetic map length and interference strength are made across strengths within sex, the strain with the higher genome wide recombinaiton rate also has stronger interference (measured as nu gamma parameter) [@ma2015_cattle]. In *Peromyscus leucopus*, the wild mice had higher gwRR and stronger (gamma nu estimates) compared to lab raised mice of the same speicies [@peterson2019].

Additionaly -- preliminary data from a previous cross using MLH1 cytology in spermatocytes in house mouse replicated these results with IFD measures from  M. m. musculus^PWD^  being longer compared to M. m. castenus^Cast^ (HVR unpublished data) (cast has low gwRR).

@@

## The CONTEXT of MEIOSIS

the study confirmed preivously identified sexual dimorhisc patterns in regard to the recombiantion landscape (in mammals) -- in the goal of unifying all of these results, we sough to investigate physiologycal aspects of gametogenesis which have conserved sexual dimorphic features. A fundamental difference in meiosis (MI) for males and females is the presence of the centrosome which serves as a nucleation site for MT at each pole (and changes the shape of the spindle). (Our hypothesis is that the tension in a centrosome spindle is stronger (or more uniform) since all MT-KT are anchored at single points across cells (the two centrosomes). This will also make prediction for the spindle assembly checkpoint (SAC), which is stronger/stricter in smaller gametes with centrosomes.  (the SAC is more sensitive to achiasmate bivalents on the spindle / better at detecting a lack of tension).
Whereas in the acentrosome spindle is more diffuse (across a larger area) and the tension across bivalents are anchored by multiple MTOCs, resulting in weaker tension force or less uniform strengths across all the bivalents within the cells. The SAC in eggs seems to be weaker/leakier, that is more achiasmate bivalents are required to trigger the SAC (stop the division).

**main sex differences in spindle** are the anchoring of MT at opposite poles

in males there are centromsomes for Mi and MII divisions
where are in oocytes, a (universal) conserved feature of oogenesis in many species is that they loose centrosomes -- instead the MTs are anchored at mulitple aMTOCs.

There is potential for these aMTOC to be unevenly distributd across the cortical and central side of the spindles (Wu et al, Schuh and ellenberg) --- (could be an origin of different tension / pulling )

The survalence of SAC in oocytes might be decreased due to large volumne difusing the signals.

#CO placement translates to bivalent shape

**still an open question of if and how the CO number and placement of crossovers translates to different fitness values for segregation rates**

the metric / amout of sister cohesion might be an good place to start

the ultimate goal is proper disjunction of chromosomes, the main checkpoint is spindle assembly checkpoint (SAC) which responds to the tension signal across chiasmata

we like this model because it can be applied to **both our results** for i) sexual dimorphism variation and ii) the fast male evolution in gwRR and recombination landscape.

Uses / centers around one of the best known selection factors on meiotic division, anneuploidy. Several other facors related to meiotic recombiniation center on **segregation** (obligate chiasmata, lower bound of the genome wide rate) it's not unbelievable that the recombination

Consequences of evolution in the number and placement of crossovers / evolution of interference strength:
Spindle / SAC model: The number and placement of crossovers changes the resulting amount of sister cohesin involved the tension signal, which affects the tetrad/chiasmata structure at MI when the homologs are pulled apart (Hollis et al 2020, Lee et al 2019).  This model predicts a difference in the meiotic spindles or division mechanism between high and low rec strains. 
ii. SACE: For the 2CO in the high rec males, larger blocks of chromosomes are kept together in the next generation.
This model would predict greater reproductive variance for males.

@@

#blah

<5. Model prediction
The models above which predict directional selection on the recombination landscape in males: SACE and Spindle based selection  are most parsimonious for the models review in this paper. The reduction principle wouldn't predict males diverge in genome wide rates and two locus modifier doesn't predict males gwRR would be higher than females.>

<3.Two-locus (protect for drivers)
This second modifer model was persented in [@brandvain2012scrambling] the basics of the model are that females will have a longer genetic map and more centromere biased recombination landscape compared to males.

A strength of this model is that is presents a clear driver / source of differential fitness (ie meiotic drivers can decrease fitness/fertility). This model also accounts for the general female pattern,

However, this model doesn't explain the borad scale pattern for the male pattern.>

<Also:
	-These results are different from the sex differences because the difference is seen in both the raw and normalized IFD, indicateing the mechanism  / physcial strength (interference strength (signal) mediated through the SC has evolved.>

<Some of the traits related to faster male evolution -- distinguish in which models can redict them: the reverseal of heterochiasmy direction and the correlated evolution of gwRR and interference strength.

**DSB evolved but, don't included as an examined trait, since most evolution based models don't make predictions for this stage of meiosis.**
_the propotion of DSB repaired into crossovers hasn't evolved, the total number of DSBs has evolved_  (following the tethered lop model of DSBs formation -- this would suggest that in the high rec strains, assuming no large difference in the Mb size of the chromosomes, 
during chromatin remodeling in prophase, in the high rec strains the chromosomes have longer axis, allowing a higher number but shorter DNA loops (for SPO11 to make the DSBs). Our results mirror those found in Baier et al [@baier2014]

Our results show that the ratios of DSB : CO 
**range from X to X which is similar to those presented in [@baier2014] (which report X to X) **

These conserved ratios suggest that the increase in DSB numbers are accompanied by proportional increase in crossovers.

This shifts support away from evolution at the CO:NCO decision to an earlier prophase stage such as when the chromatin compaction is set up. **increase in DSBs is driven by the number of DNA loops** and due to the inverse relationship of loop and chromatin axis length, this would result in longer sc lengths (which reflect the underlying chromatin axis length).>

<Petkov paper -- states that the sexual dimorphism for interference strength is due to the chromatin structure (meiotic chromosome organization) with the increased interference strength in males being due to longer chromatin loop lengths.

same pattern in humans (Gruhn et al 2013)
  (exceptions, birds (fish?) plants?) -- examining this is hard due to chromosome size effecs
  - This also assumes that the DNA loop sizes have uniform size along and between chromosomes

2. Short axis / longer DNA loops will have stronger interference
This prediction combines the findings of chromatin compaction and interference differences which are intertwined by the transmission of interference along the chromosome axis and the inverse relationship chromatin compaction (axis length and DNA loop sizes).
In physical units (micron) the counting model fits interference well to the 2D linear space of the axis these measures of interference are conserved across sexes in mouse. This is because the counting units are the BASES of DNA loops which is conserved along the 2D length of chromatin axis. When this space is transformed to account for all of the DNA base pairs which are sequestered into the DNA loops above the axis, the strength of interference will vary depending the on underlying chromatin structure.>

<We focused on the largest axis of variation, the faster evolution of genome wide recombination rates in males. One of our primary objectives was to identify other traits which correlate (could explain the increase).>

<Baier et al found significant differences in sc legnths for 
The tests in our study - show moderate to low significance in the differences in SC lengths. (this is because we have more noise-can't distinguish individual chromosomes)

a study of in house mouse -- using different markers for DSBs and intermediates (markers for the recombination pathway (from DSB to mature crossover))
-different DSB marks, distinct inbred strains two classical lab strains and a wild derived strain from castenus, used FISH for chromosome specific measures

our results are similar in several ways
- correlation of final MLH1 and DBS counts, SC measues correlate with the DSB number (even tho the differnces we observe in our study are -- slight / less significant -- we attribute that to chromosome specific noise )>

<We can't distinguish if this is 3 independent instances of evolution in gwRR OR if the patterns are due to shared standing variation, incomplete lineage sorting, but these points are for describing the general patterns.>
<extends the lenomand focusing on two conditions. first that the stength of epistatsis in dependant on the cis or trans nature. the authors note that this would apply to gene archtechture (and increase with in cis ) -- where the regulatroy region and coding region of a gene evolve (together).  second that the relative interference strengths differnt between male and female.  This can apply to (genes acting in sexual antagonism) or can be mediated via imprinted genes. 
originally proposed/outlined by Trivers [] and refined /formalized in [@lenormand2003]. This model describes a system where gametes are competing or selection is acting in the haploid stage to cause difference in genome wide recombination rates between males and females. This is a modifier model of three loci, where the recombination rate between two loci is determined by a third locus (or modifier)>
<2.  Review the assumptions for measuring CO interference in different ways.
The logical model: Crossover interference is a mechanism of suppressing crossovers, negative relationship between interference strength and gwRR. More crossovers can fit along the chromosomes and results in more crossovers overall.
Examples of empirical results supporting this pattern;  Otto Payseur, (Ruiz-Herrera? Bomblies?)
Quantifing using the gamma distribution: 
Caveats,  This doesn't make since for the general recombination landscape:   i) finite chromosome length and ii) limited range of crossovers per chromosomes (1 to 4).
Alternative interference characterization:  i) the number of crossovers are held constant and ii) transform / generalize the rec landscape, (focus on how the rec landscape would change instead of the genome wide rate) iii) assume crossover assurance.
	-The Goldstein model, models predicts a positive correlation with interference strength and higher recombination rate. A modifier which changes the probability of the recombination landscape having 0 crossovers to 2 crossovers, (driven by crossover assurance)>

>>>>

---
title: "Discussion Draft"
date: ''
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
  pdf_document: default
  word_document:
    toc: yes
    toc_depth: '5'
bibliography: refs_Main_Report.bib
---

@@

#reversed hetC direction

Reversed heterochaiasmy (briefly noted in the first part)
The common direction of heterochiasmy is female biased, however male biased heterochiasmy has (significant number of observations.
-- has been seen before (cattle, marsupial)-- Our results are novel, because the pattern is indicative of rapid evolution (male specific) -- polymorphism of low and high group.

models
the prediction from two of the modifier models, haploid selection and SACE are that males will generally evolve lower recombination rates overall compared to females. (cases with exceptions.? we don't have evidence or reason to predict these exception cases are held)
The two locus modifier model  does make predictions / parameterize cases for (reversal of the common pattern of female biased heterochiasmy). Which are dependant on the stage of action for the driver and the linkage status (see table 1 Brandvain and Coop).
There are no prediction for reversal of heterochiasmy direction  under the COM (pairing model, but it should be noted that this model wasn't designed to incorporate this type of data.

Under the spindle based model faster male specific evolution of the genome wide recombination rate is primarily driven by differences in the strength of the SAC.
 leading to different evolutionary trajectories for the genome wide recombination rate.

The SAC 'read's that all tetrads have tension -- (each set of centromeres have sinlge and opposite pole connection (synthellic?) and that the sister cohesion connecting the homologs -- sets up the tension -- opposes the pull from each of the poles.

Efficient normal SAC -- works by pausing entry into anaphase or triggers apoptosis -- if a unpaired / tension-lacking tetrad is detected
(sperm SAC is finer tuned (can detect 1-2 tetrads lacking tension -- ), while the female SAC  (has a higher treshold for being triggered  (4-5?)
(Land Kauppi,

(relaxed selection - / inefficient SAC -- in females -- fits with higher rate of achiasmata bivalents (there might be alternative mechanisms facilitating achiasmate segregation)

  Mechanisms for sex differences in the strength of the SAC are the cell volumne (diffusion of signal molecules) and the centrosome being present at the spindle have been proposed in the literature.
(testing / proving the mechanism for the new optima is for the high recombining group in males is out of the scope of this paper, but we propose some hypothesis are suggested below.
 but the evolutionary pattern fits that of directional.

@@

Our results on the broad scale pattern of crossover placement are generally supported by the literature. As reviewed in (Sardel Kirkpatrick), sexual dirmorohism in the broad scale landscape is a highly conserved trait.  However there are exceptions, (marsupial, bird?)
the popgen models
The two locus modifier model and SACE model make predictions for (sexual dimorphism ) in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks respectively.  
functional models
The two mechanical models also both predict sexual dimorphism in the recombination landscape for distinct reasons. The chromosome pairing based model (COM), predicts the sex differences in the positioning of crossovers is due to a combination of the length of the axis and intensity of the RPMs.  The spindle based model proposed that sexual dimorphism in the recombination landscape is driven by difference strength in the spindle assembly checkpoint (Lane Kauppi). The relaxed selection imposed by an ineffective SAC results in a uniform distribution of positions (while maintaining positive interference). Whereas,  in males, the directional selection could cause decreased variance in the recombination landscape via telomeric positioning of crossovers. Under the telomeric bias  recombination landscape, entry into anaphase may be more synchoriznes (and faster) given that the reductional separation of homologs is dependent on a minimal amount of sister cohesion connecting the homologs (Figure X).

[@sardell_sex_2020].

@@

**the within animal variation is equvillent to between cell variation**
[@lynn2002]

Our results of greater between cell variation . these results are supported by human and mouse (Gruhn (Kong)?  (Zwel -- Cole).  These data are also scant, between cell variance isn't often measured especially for both sexes. (given they require single cell observations for both sexes).

While the modifier models parametrize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant. (selection on penetrance).
In the spindle based selection model,  the source of between cell variance is attributed to difference strength of checkpoints. (Sexual dimorphism in the stringency of the spindle assembly checkpoint is a well established result in the literature. Most empirical results come from mammals (is this true?) yet, some of the hypotheses for the mechanism of inefficient SAC in oocytes are attributed to acentrosomal spindles (cite) and cell volume (cite), which are conserved sexual dirmohpic features of gametogenesis (cite).

# Chromatin Structure

@@

The results of sexual dimorphism in chromatin structure is supported by the literature, especially for mammals (Lynn, Gruhn, ). 
models
Very few models make predictions for how chromatin organization relates to variation in recombination rates / the recombination landscape.  (the chromatin organization is how chromosomes with the same number of base pairs) are transformed into bottle brush structures with a central axis and most of the genetic material arranged as loops emanating outwards (figure).
Only the COM model states that chromatin structure will be longer in oocytes due to greater cell volume.  This prediction / model might fit broader pattern (such as in At where pollen is the larger cell and has longer axis length) (cahoon libuda).  
Predictions for this trait are  inconclusive (due to lack of cytology measures from both sexes. We argue that the chromatin structure (and axis length) should be considered as an integral parameter when studying recombination landscape.  The sex differences in chromatin strucutre are closely connected to sex differences  in the interference strength and our results mirror those in Petkov et al.  (prediction, shorter-loop will have stronger interference).

@@

@@

@@

[@ottoPaysuer2019].

@@

[@ma2015_cattle]. In *Peromyscus leucopus*, the wild mice had higher gwRR and stronger (gamma nu estimates) compared to lab raised mice of the same speicies [@peterson2019].

Some basics of the constraints / bounds of interference are known, yet there are still many unknowns regarding the relationship of genome wide recombination rate and interference strength. 
 The non-random spacing of crossovers along the 2D length of chromosomes is the first level of proof /meeting the rejection of the (null model of random placement (cite gamma COI papers).
The physical limit of how far apart crossovers can be placed on a chromosome is a function of the length of the meiotic chromosome axis.
Logically a negative correlation is expected; increasing the number of crossovers across chromosomes would most logically be done by  more densely spacing crossovers along chromosomes and decreasing interference strength. This pattern has empirical support -- and describes  -- the fundamental relationship / of SC area of crossover number which is observed in much of the literature (which describes large/broad scale patterns).

Our results support an opposite pattern, a positive correlation between interference strength and genome wide recombination rates; the strains which show rapid evolution of the gwRR also show stronger interference strength.  We've found a number of examples of positive correlation between genome wide recombination rate and interference. The within sex comparison of two breeds of cattle (Ma et al), the Peromyscus species in (Peterson et al ), and preliminary data (HVR un published).

Theoretical models haven't really considered the evolution of interference strength -- While the COM model predicts that interference (and recombination landscape) arises from known oscillatory movements during prophase it lacks an evolutionary based prediction.
While the SACE modifier model does not explicitly model evolution of interference strength -- we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a landscape with stronger interference strength.
Veller et al rbar -- and predictions / for this - found a similar pattern (positive correlation in their simulations of male human data.

Models from Goldstein et al (review in Otto and Payseur) suggest that if this pattern is widespread --- interference evolves in X manner.

@@

@@
